Generalized measurement invariance tests with application to factor analysis

### 2011-12-13 www/paper.Rnw
Modified example section:   - Revised text for simplicity   - Modified R code to work with lavaan-S3.R, as opposed to wdhfit objects.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@84 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\citeA{WicDol05} utilized confirmatory
factor models to study measurement invariance in a series of
general intelligence scales.  They were interested in the
notion of {\em{stereotype threat}}, whereby stereotypes concerning a
subgroup's ability adversely impact the subgroup's performance on tests of that
ability.  The authors specifically focused on stereotypes concerning
ethnic minorities' performance on tests of general intelligence.  In this section, we used the proposed tests to supplement the authors' original analyses.

@@

To study stereotype threat in a measurement invariance framework, Study 1 of
\citeA{WicDol05} involved 295 high school students
completing three intelligence tests in two between-subjects conditions. 
Conditions were defined by whether or not students received primes about ethnic
stereotypes prior to testing. To study the data, Wicherts et al.\ employed a
four-group, one-factor model with the three intelligence tests as manifest
variables. The groups were defined by ethnicity (majority/minority) and by the
experimental manipulation (received/did not receive stereotype prime). Results
indicated that the intelligence tests lacked measurement invariance, with the
minority group receiving stereotype primes being particularly different from the
other three groups on the most difficult intelligence test.  Most notably, the
factor loading and intercept for the ``minority, stereotype prime'' group were
smaller, and the unique variance was larger.

As Wicherts et al. describe, one explanation for this pattern of results involves the idea that ability is positively related to ``domain identification''
\cite{Ste97}: individuals who are high on ability may be more prone to stereotype threat because they are more ``tied in'' to the domain of interest and are
more aware of stereotypes.  To test this idea, we employ a four-group, confirmatory factor model used by Wicherts et al.\ in tandem with the tests proposed
in this paper.  Our $V$ variable is participants' grade point averages (GPA, which were unused in the original analyses), which will reflect participants' identification with the academic domain.

\subsection{Method}
Wicherts et al.\ tested a series of confirmatory factor models for
various types of measurement invariance.  We focus on the model used
in their Step~5b (see pp.~703--704 of their paper), which involved
across-group parameter restrictions on the factor loadings, unique
variances, intercepts, and factor variances.  These parameters were
typically restricted to be equal across groups, though a subset of the
parameters were allowed to be group-specific upon examination of
modification indices.  Most notably, the most difficult intelligence test's factor loading, unique variance, and intercept differed for the ``minority,
stereotype threat'' group, as compared to the other groups.
Factor means were also estimated for three of the groups.  The model
provided a reasonable fit to the data, as judged by examination of
$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
(comparative fit index) statistics.  Wicherts et al.\ also had a
Step~6 (the final step), where they examined a model that was simpler than
described above and that provided a slightly better fit to the data.
This model restricted factor means to be equal across a subset of the
groups, however, which was not helpful for our purposes (because we are focusing on a single group).

The stereotype threat effect of interest is most relevant to the
"minority, stereotype prime" group.  Thus, to test the idea that
domain identification impacts the magnitude of stereotype threat
effect, we focus on parameters specific to this group.  We test these
parameters for measurement invariance across the GPA variable, with
violations supporting the hypothesis that domain identification
impacts the magnitude of the stereotype threat effect.  We also test
the intercept parameter individually.  This is because, if domain
identification does impact stereotype threat, it could be
conceptualized as the error term having a nonzero expectation across
individuals (i.e., individuals with high domain identification have a
negative error expectation and individuals with low domain
identification have a positive error expectation).  The expected value
for the error term cannot be uniquely separated from the intercept, so
that the estimated intercept would display a lack of measurement
invariance (see, e.g., \citeNP{Mil11}, p.\ 77).

To carry out
the tests, the four-group confirmatory factor model described above
was first fit to the data.  Casewise derivatives and 
the observed information matrix were then obtained, and they were used
to calculate the cumulative score process via
\eqref{eq:cumscore}.  Finally, we
obtained various test statistics and $p$-values from the cumulative score
process.  These include the double-max statistic from \eqref{eq:dmax},
the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
statistic from \eqref{eq:maxlm}.

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change.
That is, we have the ability to 
test various subsets of 
parameters.  For the Wicherts et al.\ data, we first test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\nu_{i,1,\text{MSP}}\ \lambda_{i,1,\text{MSP}}\
\theta_{i,1,\text{MSP}}\ \alpha_{i,1,\text{MSP}}) = 
(\nu_{0,1,\text{MSP}}\ \lambda_{0,1,\text{MSP}}\
\theta_{0,1,\text{MSP}}\ \alpha_{0,1,\text{MSP}}),\ i=1,\ldots,n_{\text{MSP}},
\end{equation}
where $\nu_{i,1,\text{MSP}}$ reflects the intercept for student $i$ on intelligence
test 1 in the ``minority, stereotype prime'' group, with other
subscripts defined similarly.  Additionally, $\lambda$ is a factor loading, $\theta$
is a unique variance, and $\alpha$ is a factor mean, and the
hypothesis is taken only wrt the $n_{\text{MSP}}$ individuals in the
``minority, stereotype prime'' group.  In words, the hypothesis states
that the four group-specific model parameters are invariant with
respect to GPA.

We can construct a similar hypothesis involving only the intercept
parameter
\begin{equation}
    \label{eq:exh0.2}
H_0:\ \nu_{i,1,\text{MSP}}\ = \nu_{0,1,\text{MSP}},\ i=1,\ldots,n_{\text{MSP}},
\end{equation}
which, as described previously, tests a more specific theory of the
manner in which domain identification impacts stereotype threat.
We consider tests of both \eqref{eq:exh0} and \eqref{eq:exh0.2}
below.

\subsection{Results}
<<mz_mod>>=
d <- dget("WichertsEtAl2005_Study1.txt")

wdh0 <- wdhfit(d)

# Refit model with new identification constraints
wdh0.1 <- wdhfit2(d)

## subsample models
## This could be used to compare sequential LRTs with the LM
## statistic that tests all parameters simultaneously.
## (This is not done in the data example.)
##############################################################
## lrstat <- function(gpa) 2 * as.vector(logLik(wdhfit(d[d$gpa <= gpa,])) + logLik(wdhfit(d[d$gpa > gpa,])) - logLik(wdh0))
## wdh.gpa <- sort(d$gpa[d$gpa >= quantile(d$gpa, 0.1) & d$gpa <= quantile(d$gpa, 0.9)])
## # Results in many "computationally singular" errors:
## wdh.lrstat <- sapply(wdh.gpa, lrstat)
## save(wdh0, wdh.gpa, wdh.lrstat, file = "wicherts_mod.rda")
@

@@

%% Details on test results
The results displayed in Figure~\ref{fig:cusumex} generally show that the
``minority, stereotype prime'' group lacks invariance with respect to
GPA.  All three tests are significant at the .05 level when the four
group-specific parameters are tested simultaneously (first column).
When the intercept parameter is tested individually, two of the three
tests achieve significance (the CvM test and $\max \mathit{LM}$
statistics).  These results generally support the notion that domain
identification (as measured by GPA) influences the magnitude of
stereotype threat.  The $k^*=1$ results also imply that the tests
differ in power and/or in error rates, an issue we address later in the
paper.

%% add paragraph about peaks of graphs
We can also use the test results to estimate the threshold $\nu$,
which in this context can be used to define GPA-based subgroups whose
measurement parameters differ.  As described previously, estimates of
$\nu$ can be obtained by 
examining the peaks in Figure~\ref{fig:cusumex}.  For the $k^*=4$
panels (first column), two peaks occur near GPAs of 6.0 and of 6.3 (6
corresponds to the American `C' range).  These peaks roughly divide
individuals into those receiving D's and F's, those receiving solid
C's, and those receiving high C's, B's, and A's.  Parameter-specific
results can also be obtained, as shown for the intercept parameter in
the second column of Figure~\ref{fig:cusumex}.  Here, a single peak
occurs in the three panels around a GPA of 7.0 (`B' range).

\begin{figure}
\caption{Three test statistics of~\eqref{eq:exh0} (set of
  group-specific parameters)
  and~\eqref{eq:exh0.2} (intercept only) across values of student GPA, 
  using data from Study 1 of Wicherts et al.\ (2005).
  Solid, horizontal lines represent critical values
  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
  represent values of the Cram\'{e}r-von Mises test statistic.}
\label{fig:cusumex}
\setkeys{Gin}{width=\textwidth}
<<gefp-4-1, fig=TRUE, height=8, width=6>>=
## add a small amount of random noise to duplicated GPAs, so that everything
## works with gefp().  (Without this, one can still obtain DM and CvM test 
##                      results but not supLM results.)
set.seed(1090)
ties <- duplicated(d$gpa)
d$gpa[ties] <- runif(sum(ties), d$gpa[ties]-.01, d$gpa[ties]+.01)
## empirical fluctuation processes (4 vs 1 par)
gefp_4_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 15:18, vcov = info.wdhfit, sandwich = FALSE)
gefp_1_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 18, vcov = info.wdhfit, sandwich = FALSE)

# processes for second identification constraints
gefp_4_info_1 <- gefp(wdh0.1, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 15:18, vcov = info.wdhfit, sandwich = FALSE)
gefp_1_info_1 <- gefp(wdh0.1, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 18, vcov = info.wdhfit, sandwich = FALSE)

@@

\subsection{Discussion}
Application of the proposed measurement invariance tests to the
\citeA{WicDol05} data allowed us to study the
domain identification's impact on stereotype threat in a
straightforward manner.  The tests focused on a set of group-specific
parameters within a four-group confirmatory factor model, and they
could be carried out using the results from a single estimated model.
The latter fact is important, as other approaches to the problem
described here may adversely impact model fit and degrees of freedom (e.g., if
GPA is inserted directly into the model) or require multiple models to
be estimated (e.g., for LRTs).  Further, through examination of both
test statistics and cumulative score processes, the tests were easily
interpretable from both a theoretical and applied standpoint.  In the
next section, we study the proposed tests' power via simulation.

>>>>

\citeA{WicDol05} utilized confirmatory
factor models to study measurement invariance in a series of
general intelligence scales.  They were interested in the
notion of {\em{stereotype threat}}, whereby stereotypes concerning a
subgroup's ability adversely impact the subgroup's performance on tests of that
ability.  The authors specifically focused on stereotypes concerning
ethnic minorities' performance on tests of general intelligence.  In this section, we use the proposed tests to supplement the authors' original analyses.

@@

To study stereotype threat in a measurement invariance framework, Study 1 of
\citeA{WicDol05} involved 295 high school students
completing three intelligence tests in two between-subjects conditions. 
Conditions were defined by whether or not students received primes about ethnic
stereotypes prior to testing. To study the data, Wicherts et al.\ employed a
four-group, one-factor model with the three intelligence tests as manifest
variables. The groups were defined by ethnicity (majority/minority) and by the
experimental manipulation (received/did not receive stereotype prime). Results
indicated that the intelligence tests lacked measurement invariance, with the
minority group receiving stereotype primes being particularly different from the
other three groups on the most difficult intelligence test.  In the
example below, we employ one of the models used by Wicherts et al.
Our $V$ variable is participants' grade point averages, which were unused in the original analyses.

\subsection{Method}
Wicherts et al.\ tested a series of confirmatory factor models for
various types of measurement invariance.  We focus on the model used
in their Step~5b (see pp.~703--704 of their paper), which involved
across-group parameter restrictions on the factor loadings, unique
variances, intercepts, and factor variances.  These parameters were
typically restricted to be equal across groups, though a subset of the
parameters were allowed to be group-specific upon examination of
modification indices.    The model
provided a reasonable fit to the data, as judged by examination of
$\chi^2$, RMSEA (root-mean-square error of approximation), and CFI
(comparative fit index) statistics.  While the model included four
groups as described above, we focus only on the submodel for the minority
group that received stereotype primes.

To carry out
the tests, we first fit the four-group model to the data,
calculating casewise derivatives and 
the observed information matrix.  We then
calculated the cumulative score process wrt student GPA via
\eqref{eq:cumscore}, allowing us to
obtain various test statistics and $p$-values from this 
process.  Test statistics include the double-max statistic from \eqref{eq:dmax},
the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the $\max \mathit{LM}$
statistic from \eqref{eq:maxlm}.

As mentioned in the theory section, the 
tests give us the flexibility to study hypotheses of partial change.
That is, we have the ability to 
test various subsets of 
parameters.  For the Wicherts et al.\ data, we first test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\nu_{i}\ \lambda_{i}\
\theta_{i}\ \alpha_{i}) = 
(\nu_{0}\ \lambda_{0}\
\theta_{0}\ \alpha_{0}),\ i=1,\ldots,n_{\text{MSP}},
\end{equation}
where $\nu_{i}$ is the intercept for student $i$ on intelligence
test 1, $\lambda_{i}$ the factor loading, $\theta_{i}$ the unique
variance, and $\alpha_i$ the factor mean.
The individuals are ordered according to GPA, and the 
hypothesis is taken only wrt the $n_{\text{MSP}}$ individuals in the
``minority, stereotype prime'' group.  Thus, \eqref{eq:exh0} states 
that these four group-specific model parameters are invariant with
respect to GPA.

We can construct a similar hypothesis involving only the intercept
parameter
\begin{equation}
    \label{eq:exh0.2}
H_0:\ \nu_{i}\ = \nu_{0},\ i=1,\ldots,n_{\text{MSP}},
\end{equation}
which, e.g., may be used to isolate the impact of stereotype threat.
We consider tests of both \eqref{eq:exh0} and \eqref{eq:exh0.2}
below.

\subsection{Results}
<<mz_mod>>=
load("wicherts05_study1.rda")

d <- subset(d, !is.na(gpa))

## fit model 5b
wdh5b <- lavaan(
   'ability   =~ label(c(rep("load_n", 3), "load_n:min_t")) * numerical + label(rep("load_v", 4)) * verbal + 1 * abstract
    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
    verbal    ~  label(rep("mean_v", 4)) * 1
    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
    abstract  ~~ label(rep("var_a", 4)) * abstract
    verbal    ~~ label(rep("var_v", 4)) * verbal',
  data = d, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")

## Refit model with new identification constraints
wdh5b.2 <- lavaan(
   'ability   =~ 1 * verbal + label(c(rep("load_n", 3), "load_n:min_t")) * numerical  + label(rep("load_a", 4)) * abstract
    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
    verbal    ~  label(rep("mean_v", 4)) * 1
    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
    abstract  ~~ label(rep("var_a", 4)) * abstract
    verbal    ~~ label(rep("var_v", 4)) * verbal',
  data = d, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
@

@@

%% Details on test results
The results displayed in Figure~\ref{fig:cusumex} generally show that the
``minority, stereotype prime'' group lacks invariance with respect to
GPA.  All three tests are significant at the .05 level, both when the four
group-specific parameters are tested simultaneously (first column) and
when the intercept parameter is tested individually (second column).

We can also use the test results to estimate the threshold $\nu$,
which in this context can be used to define GPA-based subgroups whose
measurement parameters differ.  As described previously, estimates of
$\nu$ can be obtained by 
examining the peaks in Figure~\ref{fig:cusumex}.  For the $k^*=4$
panels (first column), two peaks occur near GPAs of 5.9 and of 6.3 (6
corresponds to the American `C' range).  These peaks roughly divide
individuals into those receiving D's and F's, those receiving solid
C's, and those receiving high C's, B's, and A's.  Parameter-specific
results can also be obtained, as shown for the intercept parameter in
the second column of Figure~\ref{fig:cusumex}.  The peak locations are
generally similar to those of all four parameters, though this finding
will not hold in general.

\begin{figure}
\caption{Three test statistics of~\eqref{eq:exh0} (set of
  group-specific parameters)
  and~\eqref{eq:exh0.2} (intercept only) across values of student GPA, 
  using data from Study 1 of Wicherts et al.\ (2005).
  Solid, horizontal lines represent critical values
  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
  represent values of the Cram\'{e}r-von Mises test statistic.}
\label{fig:cusumex}
\setkeys{Gin}{width=\textwidth}
<<gefp-4-1, fig=TRUE, height=8, width=6>>=
## add a small amount of random noise to duplicated GPAs, so that everything
## works with gefp().  (Without this, one can still obtain DM and CvM test 
##                      results but not supLM results.)
set.seed(1090)
ties <- duplicated(d$gpa)
d$gpa[ties] <- runif(sum(ties), min = d$gpa[ties] - .001, max = d$gpa[ties] + .001)

## FIXME It seems like info_full should go in lavaan_S3.R.
info_full <- function(x, ...) solve(vcov(x) * nobs(x))
scores_min_t <- function(x, ...) {
  ef <- estfun(x, ...)
  ef[d$group == "exp:minority", ]
}
gpa_min_t <- subset(d, group == "exp:minority")$gpa

## empirical fluctuation processes (4 vs 1 par)
gefp_4_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
gefp_1_info <- gefp(wdh5b, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)

# processes for second identification constraints
gefp_4_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
gefp_1_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)

@@

\subsection{Discussion}
Application of the proposed measurement invariance tests to the
\citeA{WicDol05} data allowed us to study the extent to which model
parameters are invariant wrt GPA in a straightforward manner.
The tests focused on a set of group-specific
parameters within a four-group confirmatory factor model, and they
could be carried out using the results from a single estimated model.
The latter fact is important, as other approaches to the problem
described here may require multiple models to
be estimated (e.g., for LRTs) or adversely impact model fit and
degrees of freedom (e.g., if GPA is inserted directly into the model).
Further, through examination of both 
test statistics and cumulative score processes, the tests were
interpretable from both a theoretical and applied standpoint.  In the
next section, we study the proposed tests' power via simulation.
